---
title: "Parks, Greenways, and Home Values in Raleigh, North Carolina"
author: "Megan Streicher"
---

## Introduction

This project looks at the relationship between access to green space and home values in Raleigh, North Carolina.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(here)
library(ggspatial)

options(tigris_use_cache = TRUE)
```

## Data

```{r}
#| label: load-data
#| message: false
#| warning: false

# Load ACS median home value data
home_value_raw = read_csv(here("data", "ACSDT5Y2024.B25077-Data.csv"))

# Load ACS median household income data
income_raw = read_csv(here("data", "ACSST5Y2024.S1901-Data.csv"))

# Load Wake County census tract boundaries
wake_tracts = tracts(
  state = "NC",
  county = "Wake",
  year = 2024,
  cb = TRUE
) |>
  st_transform(32119)
```

```{r}
# Check the first few rows of each dataset
head(home_value_raw)
head(income_raw)

# Check column names so we can identify the correct variables
names(home_value_raw)
names(income_raw)
```

## 3. Clean ACS data

This keeps only the columns we need: census tract ID, median home value, and median household income.

```{r}
home_values_clean = home_value_raw |>
  filter(GEO_ID != "Geography") |>
  mutate(
    GEOID = str_replace(GEO_ID, "1400000US", ""),
    median_home_value = as.numeric(B25077_001E)
  ) |>
  filter(!is.na(median_home_value)) |>
  select(GEOID, NAME, median_home_value)

income_clean = income_raw |>
  filter(GEO_ID != "Geography") |>
  mutate(
    GEOID = str_replace(GEO_ID, "1400000US", ""),
    median_household_income = as.numeric(S1901_C01_012E)
  ) |>
  filter(!is.na(median_household_income)) |>
  select(GEOID, median_household_income)

acs_data = home_values_clean |>
  left_join(income_clean, by = "GEOID")
```

## 4. Load census tract boundaries

This gets Wake County census tract shapes so we can make maps and connect the ACS data to geography.

```{r}
wake_tracts = tracts(
  state = "NC",
  county = "Wake",
  year = 2024
)

wake_tracts = wake_tracts |>
  st_transform(2264)

raleigh_tracts = wake_tracts |>
  left_join(acs_data, by = "GEOID")
```

## 5. Load Raleigh parks data

Your parks CSV is separated by semicolons, so this uses `read.csv()` with `sep = ";"`.

```{r}
parks = read.csv(
  here("data", "Parks_-7279106479596582801.csv"),
  sep = ";"
)

head(parks)
```

## 6. Clean parks data and convert to spatial data

```{r}
parks_clean = parks |>
  filter(!is.na(Point...Latitude)) |>
  filter(!is.na(Point...Longitude)) |>
  mutate(
    latitude = as.numeric(Point...Latitude),
    longitude = as.numeric(Point...Longitude),
    acres = as.numeric(Map.Acres)
  )

parks_sf = st_as_sf(
  parks_clean,
  coords = c("longitude", "latitude"),
  crs = 4326
)

parks_sf = parks_sf |>
  st_transform(2264)
```

------------------------------------------------------------------------

## 7. Create a 1-mile park access measure

This creates a 1-mile buffer around each park. Then it checks whether each census tract intersects a park buffer.

This is easier than measuring exact distance and uses the same type of spatial method from class.

```{r}
parks_buffer = parks_sf |>
  st_buffer(5280)

raleigh_tracts$near_park = apply(
  st_intersects(raleigh_tracts, parks_buffer),
  1,
  any
)

raleigh_tracts = raleigh_tracts |>
  mutate(
    near_park_text = case_match(
      near_park,
      TRUE ~ "Within 1 mile of park",
      FALSE ~ "More than 1 mile from park"
    )
  )
```

## 8. Check the final analysis data

```{r}

analysis_data = raleigh_tracts |>
  st_drop_geometry() |>
  select(
    GEOID,
    NAME.x,
    median_home_value,
    median_household_income,
    near_park,
    near_park_text
  ) |>
  filter(!is.na(median_home_value)) |>
  filter(!is.na(median_household_income))

head(analysis_data)

nrow(analysis_data)

```

## 9. Descriptive statistics table

```{r}
#| label: descriptive-table

summary_table = analysis_data |>
  group_by(near_park_text) |>
  summarize(
    tract_count = n(),
    average_home_value = mean(median_home_value, na.rm = TRUE),
    average_income = mean(median_household_income, na.rm = TRUE)
  )

summary_table
```

## 10. Map of median home values

```{r}
#| label: home-value-map
#| fig-cap: "Median home values by census tract in Wake County, North Carolina."

ggplot() +
  layer_spatial(raleigh_tracts, aes(fill = median_home_value)) +
  layer_spatial(parks_sf) +
  annotation_scale() +
  scale_fill_binned() +
  labs(
    title = "Median Home Values and Raleigh Parks",
    fill = "Median Home Value"
  ) +
  theme_minimal()
```

## 11. Map of tracts within 1 mile of a park

```{r}
#| label: park-access-map
#| fig-cap: "Census tracts classified by whether they are within one mile of a Raleigh park."

ggplot() +
  layer_spatial(raleigh_tracts, aes(fill = near_park_text)) +
  layer_spatial(parks_sf) +
  annotation_scale() +
  labs(
    title = "Access to Raleigh Parks",
    fill = "Park Access"
  ) +
  theme_minimal()
```

## 12. Bar chart comparing home values by park access

```{r}
#| label: park-access-bar-chart
#| fig-cap: "Average median home value by park access category."

ggplot(summary_table, aes(x = near_park_text, y = average_home_value)) +
  geom_col() +
  labs(
    title = "Average Home Value by Park Access",
    x = "Park Access",
    y = "Average Median Home Value"
  ) +
  theme_minimal()
```

## 13. Scatterplot of income and home value

```{r}
#| label: income-home-value-scatter
#| fig-cap: "Relationship between median household income and median home value."

ggplot(analysis_data, aes(x = median_household_income, y = median_home_value)) +
  geom_point() +
  labs(
    title = "Median Income and Median Home Value",
    x = "Median Household Income",
    y = "Median Home Value"
  ) +
  theme_minimal()
```

## 14. Regression models

Model 1 looks at the relationship between park access and home value.

Model 2 adds median household income as a control.

```{r}
#| label: regression-models

model1 = lm(
  median_home_value ~ near_park,
  data = analysis_data
)

model2 = lm(
  median_home_value ~ near_park + median_household_income,
  data = analysis_data
)

summary(model1)

summary(model2)
```

## 15. Simple regression table

```{r}
#| label: regression-table

model1_results = as.data.frame(summary(model1)$coefficients)

model2_results = as.data.frame(summary(model2)$coefficients)

model1_results

model2_results
```

## 16. Optional: income by park access

This helps you explain whether park access overlaps with wealthier areas.

```{r}
#| label: income-by-park-access

ggplot(summary_table, aes(x = near_park_text, y = average_income)) +
  geom_col() +
  labs(
    title = "Average Income by Park Access",
    x = "Park Access",
    y = "Average Median Household Income"
  ) +
  theme_minimal()
```

## 17. Short conclusion code chunk

This is not required, but it helps you quickly see your key results.

```{r}
#| label: key-results

summary_table

summary(model1)

summary(model2)
```

### Note

This project examines whether census tracts near Raleigh parks tend to have higher home values. The results show an association, not a causal relationship, because other neighborhood factors such as income, school quality, housing age, zoning, and location may also affect home values.
