---
title: "Parks, Greenways, and Home Values in Raleigh, North Carolina"
author: "Megan Streicher"
---
## Introduction

This project looks at the relationship between access to green space and home values in Raleigh, North Carolina.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(sf)
library(tidycensus)
library(tigris)
library(here)

options(tigris_use_cache = TRUE)
```

## Data
```{r}
#| label: load-data
#| message: false
#| warning: false

# Load ACS median home value data
home_value_raw = read_csv(here("data", "ACSDT5Y2024.B25077-Data.csv"))

# Load ACS median household income data
income_raw = read_csv(here("data", "ACSST5Y2024.S1901-Data.csv"))

# Load Wake County census tract boundaries
wake_tracts = tracts(
  state = "NC",
  county = "Wake",
  year = 2024,
  cb = TRUE
) |>
  st_transform(32119)
```

```{r}
#| label: check-data

# Check the first few rows of each dataset
head(home_value_raw)
head(income_raw)

# Check column names so we can identify the correct variables
names(home_value_raw)
names(income_raw)
```
## 3. Clean ACS data

This keeps only the columns we need: census tract ID, median home value, and median household income.


```{r}
#| label: clean-acs-data

home_values_clean = home_value_raw |>
  filter(GEO_ID != "Geography") |>
  mutate(
    GEOID = str_replace(GEO_ID, "1400000US", ""),
    median_home_value = as.numeric(B25077_001E)
  ) |>
  filter(!is.na(median_home_value)) |>
  select(GEOID, NAME, median_home_value)

income_clean = income_raw |>
  filter(GEO_ID != "Geography") |>
  mutate(
    GEOID = str_replace(GEO_ID, "1400000US", ""),
    median_household_income = as.numeric(S1901_C01_012E)
  ) |>
  filter(!is.na(median_household_income)) |>
  select(GEOID, median_household_income)

acs_data = home_values_clean |>
  left_join(income_clean, by = "GEOID")
```

## 4. Load census tract boundaries

This gets Wake County census tract shapes so we can make maps and connect the ACS data to geography.

```{r}
#| label: load-tracts

wake_tracts = tracts(
  state = "NC",
  county = "Wake",
  year = 2024
)

wake_tracts = wake_tracts |>
  st_transform(2264)

raleigh_tracts = wake_tracts |>
  left_join(acs_data, by = "GEOID")
```
## 5. Load Raleigh parks data

Your parks CSV is separated by semicolons, so this uses `read.csv()` with `sep = ";"`.

```r
```{r}
#| label: load-parks

parks = read.csv(
  here("data", "Parks_-7279106479596582801.csv"),
  sep = ";"
)

head(parks)
```




```{r}
```